---
title: "ArchaeoGLOBE trend analysis"
author: "Nick Gauthier and Ben Marwick"
date: "Last knit on: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document: 
    highlight: pygments
    latex_engine: lualatex
  html_document: 
    highlight: pygments
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      cache = FALSE)
```

Analysis code for the ArchaeoGlobe database. Here we use Generalized Additive Models (GAMs), a flexible form of nonlinear regression model capable of fitting smooth, time-varying trends to the ordered categorical ArchaeoGLOBE response data.

We model ordered categorical data using a latent variable following a logistic distribution. The model identifies a series of cut points, which correspond the the probabilities of the latent variable falling within each of our categories.

We fit two sets of trends. One trend is fitted to all the data simultaneously, representing the global trend across all archaeological regions. Then we fit region-level trends, which represent the deviation of each region from the global trend. By penalizing the "wiggliness" of the trend lines, we allow regional trends that don't significantly deviate from the global trend to be penalized to 0, effectively reducing that particular region to the global trend. This is a form of partial pooling, allowing the model to share information between groups and in so doing make the results less sensitive to regions with exceptionally low response rates.

After fitting the model, we can extract the region-specific trends, use a k-means clustering algorithm to group together regions with similar trends, and map the results. We repeat this analysis for both self-reported expertise and perceived data quality.

# Setup

Import packages needed for analysis. We'll use packages from the `tidyverse`, such as `readr`, `dplyr`, and `ggplot2` for data import, processing, and plotting. We'll also use `mgcv` for fitting nonlinear trends to the data. We'll use the `sf` package to help us plot shapefiles in a tidy context. The `dataverse` package allows us to pull the raw survey data from its online repository. Finally, we'll use `patchwork` (installed from GitHub) to combine multiple ggplots in the same image.

```{r, message = FALSE}
library(tidyverse)
library(mgcv)
library(sf)
library(ggplot2)
library(dataverse)

#install patchwork from github 
#devtools::install_github('thomasp85/patchwork')
library(patchwork)
```

## Data import

Read in the latest version of the ArchaeoGLOBE database and the consensus assessment from the Dataverse repository.

```{r dataverse-database, echo = TRUE}
Sys.setenv('DATAVERSE_SERVER' = 'dataverse.harvard.edu')

# get data frame of files on dataverse
ArchaeoGLOBE_Data_DOI <- 'doi:10.7910/DVN/CNCANQ'
ArchaeoGLOBE_Data_df <- get_dataset(ArchaeoGLOBE_Data_DOI)

# Only download the file we need here
ArchaeoGLOBE_Data_df_files <- ArchaeoGLOBE_Data_df$files[
  grepl('ARCHAEOGLOBE_PUBLIC_DATA|ARCHAEOGLOBE_CONSENSUS_ASSESSMENT', 
        ArchaeoGLOBE_Data_df$files$filename), ]

# read into local dir
walk(ArchaeoGLOBE_Data_df_files$label,
     ~get_file(.x, ArchaeoGLOBE_Data_DOI) %>% 
       writeBin(paste0('data/raw-data/', .x)))

# read into the current environment
archaeoglobe <- read_csv('data/raw-data/ARCHAEOGLOBE_PUBLIC_DATA.tab')
consensus <- read_csv('data/raw-data/ARCHAEOGLOBE_CONSENSUS_ASSESSMENT.tab')
```

Repeat for the archaeological regions shapefile.

```{r dataverse-shapefile, echo = TRUE, eval = FALSE}
# repeat for shapefile
ArchaeoGLOBE_Regions_DOI <- 'doi:10.7910/DVN/CQWUBI'

# get data frame of files on DV
ArchaeoGLOBE_Regions_df <- get_dataset(ArchaeoGLOBE_Regions_DOI) 

# just download the shapefile we want
ArchaeoGLOBE_Regions_df_files <- ArchaeoGLOBE_Regions_df$files[
  ArchaeoGLOBE_Regions_df$files$filename == 'ArchaeGLOBE_Regions.zip', ]

# read into local dir
walk(ArchaeoGLOBE_Regions_df_files$label,
    ~get_file(.x, ArchaeoGLOBE_Regions_DOI) %>% 
      writeBin(paste0('data/raw-data/', .x)))

unzip('data/raw-data/ArchaeGLOBE_Regions.zip', 
      overwrite = TRUE,
      exdir = 'data/raw-data/ArchaeGLOBE_Regions')

# read into the current environment, and simplify the polygons for faster plotting
regions <- st_read('data/raw-data/ArchaeGLOBE_Regions/ArchaeGLOBE_Regions.shp', 
          quiet = TRUE)
```

```{r}
# temporary workaround
regions <- st_read('../Simplified_Regions2/Simplified_Regions2.shp', quiet = TRUE) %>%
```


## Exploratory plots

Before running any analyses, let's look at the data. How many responses do we have per region?

```{r}
response_counts <- archaeoglobe %>% 
  group_by(REGION_ID) %>%
  count %>%
  mutate(n10 = replace(n, n > 10, 10))
```

```{r, fig.width = 6.5, fig.height = 5.5, echo = FALSE}
a <- regions %>%
  left_join(response_counts, by = c('Archaeo_ID' = 'REGION_ID')) %>%
  mutate(region_label = replace(Archaeo_RG, !(Archaeo_RG %in% c('Hawaii','Polynesia','Micronesia','Melanesia')), NA)) %>%
  ggplot() +
  geom_sf(aes(fill = n10), size = .3, color = 'white') +
  scale_fill_viridis_c(name = 'Contributors', guide = 'legend', breaks = seq(2, 10, 2), labels = c(seq(2, 8, 2), '10+')) +
  geom_sf_text(aes(label = region_label), nudge_x = 500000, size = 2.5, hjust = 0, expand = TRUE) +
  coord_sf(datum = NA) +
  theme_void() +
  theme(legend.position='bottom')

b <- ggplot(response_counts, aes(n, group = n)) +
  geom_bar(aes(fill = n10)) +
  scale_fill_viridis_c(guide = 'none') +
  theme_classic() +
  labs(x = 'Contributors', y = 'Regions')

a + b + plot_layout(ncol = 1, heights = c(2,1)) + plot_annotation(tag_levels = 'A')
ggsave('figures/response_distribution.png', width = 6.5, height = 5.5)
```

How many published archaeolgical excavations are estimated for each region?

```{r}
# a function for calculating the mode, from https://stackoverflow.com/a/46846474
calculate_mode <- function(x) {
  uniqx <- unique(x)
  uniqx[which.max(tabulate(match(x, uniqx)))]
}

# use this vector to order the RN_SITES variable by increasing number of sites
site_order <- c('< 50', '50-249', '250-499', '500-999', '> 1000')

# find the modal response for the number of published excavations in each region
site_counts <- archaeoglobe %>%
  select(REGION_ID, RN_SITES) %>%
  group_by(REGION_ID) %>%
  summarise(sites = calculate_mode(RN_SITES)) %>%
  mutate(sites = ordered(sites, levels = site_order))
```

```{r echo = FALSE}
sites_plot <- regions %>%
  left_join(site_counts, by = c('Archaeo_ID' = 'REGION_ID')) %>%
  mutate(region_label = replace(Archaeo_RG, !(Archaeo_RG %in% c('Hawaii','Polynesia','Micronesia','Melanesia')), NA)) %>%
  ggplot() +
  geom_sf(aes(fill = sites), size = .3, color = 'white') +
  scale_fill_viridis_d(name = 'Published Excavations', guide = 'legend') +
  geom_sf_text(aes(label = region_label), nudge_x = 500000, size = 2.5, hjust = 0, expand = TRUE) +
  coord_sf(datum = NA) +
  theme_void() +
  theme(legend.position='bottom')

sites_plot
```



## Analysis functions

Define some analysis functions that we'll be using repeatedly in the analysis, so that we don't have to keep copying and pasting the same lines of code.

This function subsets the data to highlight a variable of interest, and converts it from a wide to a long "tidy" format to make analysis and plotting easier.

```{r}
preprocess <- function(prefix, categories){
  archaeoglobe %>% # start with the full ArcheoGlobe data
    # drop columns not related to the variable of interest
    select(c(CONTRIBUTR:LAND_AREA, starts_with(prefix))) %>%
    gather(time, value, starts_with(prefix)) %>% # one value per row
    mutate(time = parse_number(time) * -1, # convert time period labels to years
           value = ordered(value, levels = categories),
           cat_num = as.numeric(value)) %>%
    mutate_if(is.character, as.factor) # convert characters to factors
}
```

This function takes a data frame produced by the above function and fits GAM to the global trend and local deviations for each region, accounting for inter-observer variability. This function takes as arguments a preprocessed data frame containing time slices, regions, contributors, and the ordered categorical response variable transformed to a numeric vector. 

```{r}
cores <- max(parallel::detectCores() / 2, 1) # physical cores for parallelization
cl <- parallel::makeCluster(cores)

fit_gam <- function(x, n_cats){
  bam(cat_num ~ 
        # this spline is for the global trend
        s(time, bs = 'cr', m = 2) + 
        # region-specific trends. bs = 'ts' and m = 1 
        # help penalize deviation from the global model
        s(time, by = REGION_LAB, bs = 'cs', m = 1) + 
        # add back in region-specific intercepts
        REGION_LAB  +
        # model contributor as a random effect
        s(CONTRIBUTR, bs = 're'),
      data = x, # data frame to analyize
      family = ocat(R = n_cats), # ordered categorical with n levels
      # final 3 arguments just speed up the model fitting
      method = 'fREML',
      discrete = TRUE,
      cluster = cl)
}
```

This function extracts the estimated trends for each region, incorporating the global and regional splines as well as the region and contributor specific intercepts. Then it clusters these trends into 6 discrete clusters using k-means. The choice of 6 clusters is somewhat arbitrary, and is made simply based on visual comparisons of different cluster solutions with the goal of ensuring visually interpretable results.

```{r}
extract_trends <-function(mod, n_clusters = 6){
  set.seed(1000) # set seed for reproducability of clusters
  archaeoglobe %>% # create dummy data for prediction in the following lines
    select(REGION_LAB) %>%
    group_by(REGION_LAB) %>%
    slice(1) %>%
    slice(rep(1:n(), each = 198)) %>%
    ungroup %>%
    mutate(time = rep_len(seq(-10000, -150, 50), n()),
           CONTRIBUTR = 'CYRBU') %>% # select an arbitrary contributor
    mutate(preds = predict(mod, .)) %>% # estimate trend lines
    mutate(preds = plogis(preds)) %>% # transform responses to [0,1] scale
    spread(time, preds) %>%
    # next is the actual kmeans clustering code
    mutate(cluster = kmeans(.[,-c(1,2)], n_clusters, iter.max = 100, nstart = 100)$cluster)
}
```

# Analysis

Now we use the functions defined above on the ArchaeoGlobe data. For convenience, first define a data frame that lists the prefixes of the variables we are interested in (e.g. "EXP" for expertise) and the levels of the ordered factors associated with each variable. This will make it easier to quickly focus on a specific variable. The `tribble` command is simply a way to make a data frame by row rather than column, which makes the code easier to read.

```{r}
response_levels <- tribble(
  ~prefix, ~categories,
  'EXP', c('None', 'Low', 'High'),
  'DQ', c('Unknown', 'Low', 'Moderate', 'Good'),
  'HUNT', c('none', 'minimal (<1%)', 'common (1-20%)', 'widespread (>20%)'),
  'EXAG', c('none', 'minimal (<1%)', 'common (1-20%)', 'widespread (>20%)'),
  'INAG', c('none', 'minimal (<1%)', 'common (1-20%)', 'widespread (>20%)'),
  'PAST', c('none', 'minimal (<1%)', 'common (1-20%)', 'widespread (>20%)'),
  'URBN', c('Absent', 'Present')
)
```

Now map each of the above functions to each variable. This allows us to run the analysis for all variables of interest in a single step, and save all the outputs in a tibble format for easy plotting. If you're running this for the first time, it should take about 40 minutes to run on a Intel NUC with a 5th-gen Intel Core i7-5557U processor and 16gb of RAM running Linux.

```{r echo = T}
# A simple type of caching...
# Do we want to run the modelling code, or
# load a previously saved result from disk, or
# download a previously saved result from a repository?
# default is not run, then check if there is a saved file, and use that, or download

# devtools::install_github('centerforopenscience/osfr')
library(osfr)

rerun_time_consuming_analysis <- FALSE # FALSE means do not run the modelling code when knitting

if(rerun_time_consuming_analysis) {
  message('running the modelling code, this may take 30-50 min...')
  # go to the next chunk of code
} else {
  # check if there is a local file and if so, load it
  if(file.exists('data/derived-data/trend_dat.rda')) {
    # the file exists on the local disk, so just read it in
    message('Loading previously saved model results from disk...')
    trend_dat <- readRDS('data/derived-data/trend_dat.rda')
  } else {
  # we don't want to run the modelling code, and the result don't exist locally,
  # so download
  message('Downloading previously saved model results, takes 2-3 min...')
  trend_dat <- osf_retrieve_file('kcr2e') %>% osf_download('data/derived-data/trend_dat.rda')
  message('Loading the data downloaded from osf.io...')
  trend_dat <- readRDS('data/derived-data/trend_dat.rda')
  writeLines(paste0('trend_dat.rda downloded from https://osf.io/kcr2e/ on ', Sys.Date()), con = 'data/derived-data/README.md', sep = '')
  message('Done.')
  } 
}
```

```{r, eval = rerun_time_consuming_analysis}
trend_dat <- response_levels %>%
  mutate(data = map2(prefix, categories, ~preprocess(.x,.y)),
         n_cats = map_dbl(categories, length), 
         mod = map2(data, n_cats, fit_gam),
         trends = map(mod, extract_trends))

# save to disk
saveRDS(trend_dat, file = 'data/derived-data/trend_dat.rda')

# write a note to indicate the provenance of this file
this_commit <-  git2r::revparse_single(git2r::repository('.'), "HEAD")

writeLines(paste0('trend_dat generated on ',
                   Sys.Date() , ' from archaeoglobe.Rmd at git commit ',
                    this_commit$sha, ' made by ',
                    this_commit$author$name, ' on ',
                    this_commit$author$when, " with the message '",
                    this_commit$summary, "'"),
           con = 'data/derived-data/README.md', 
           sep = '')
message('Done.')
```

```{r eval = FALSE, echo = FALSE}
# If we have freshly produced model output, we want to 
# upload to osf.io, this needs to be done manually because we need to 
# authenticate and don't want to put our PAT in the Rmd

osf_auth('my-PAT-for-OSF') # Navigate to https://osf.io/settings/tokens/

# upload the model data
osf_retrieve_node('gcfs6') %>% 
  osf_upload(path = 'data/derived-data/trend_dat.rda', 
             overwrite = TRUE)

# upload the README
osf_retrieve_node('gcfs6') %>% 
  osf_upload(path = 'data/derived-data/README.md', 
             overwrite = TRUE)
```

# Results

First we plot out the global trends for each land use type. Then we plot out the local (regional trends) for all land use types, and map out their associated clusters. Please refer to the .rmd source file for the code to make the plots.

The global trend in foraging shows constant high prevalence until around 6,000 years ago, after which there is a smooth decline until the present day when it is very rare. Mapping out the clusters reveals a clear east-west divide, which regions in Afro-eurasia seeing foraging earlier then the global mean, and regions in the Americas and Oceania seeing later peaks in foraging.

The global trends in the prevalence of pastoralism, extensive and intensive agriculture, and urbanism all follow a sigmoidal curve, which means the trend is linear on the scale of the linear predictor (the ordered categorical GAM uses a logit transform as a latent link function). This means that there is a simple increase in the probability of each land use type being prevalent over time.

How does self-professed level of expertise vary in each region over time? The global trend is a roughly linear increase in self-reported expertise from 10ka BP up to 2ka BP, then a falloff continuing to the present day. The present day expertise values are approximately the same as at 10ka BP. This makes sense, as it points to both the increased frequency of preserved archaeological materials with time as well as the reduction in archaeological attention in periods with extensive historical records.

The global trend in data quality is more or less the same as the expertise data, with the peak in data quality occurring more recently than for expertise and with a less dramatic falloff leading to the present day. Unlike expertise, which reaches the same values at 10ky BP and present, data quality in the present day remains high in spite of the falloff in the last 2 millennia. Also note the confidence interval for the global trend is generally wider than for the expertise responses.

## Global Trends

```{r echo = FALSE, fig.width = 4.5, fig.height = 6}
cutpoints_dat <- trend_dat %>%
  select(prefix, mod) %>%
  mutate(cutpoints = map(mod, ~.$family$getTheta(TRUE))) %>%
  select(-mod) %>%
  mutate(type = recode_factor(prefix,
                              'EXAG' = 'Extensive Agriculture',
                              'INAG' = 'Intensive Agriculture',
                              'PAST' = 'Pastoralism',
                              'HUNT' = 'Foraging',
                              'URBN' = 'Urban',
                              'EXP' = 'Expertise',
                              'DQ' = 'Data Quality')) %>%
  filter(prefix %in% c('HUNT','PAST','EXAG','INAG')) %>%
    mutate(
        cutpoints = map_chr(cutpoints, paste, collapse = ",")
    ) %>% 
    tidyr::separate(
        cutpoints, 
        c('Minimal', 'Common', 'Widespread'),
        ",", convert = TRUE
    )

global_trends <- trend_dat %>%
  select(prefix, mod) %>%
  mutate(trends = map(mod, ~plot.gam(., select = 0)[1]),
         trends = map(trends, ~.[[1]]),
         trends = map(trends, ~tibble(time = .$x, fit = c(.$fit), se = .$se))) %>%
  select(-mod) %>%
  unnest %>%
  mutate(type = recode_factor(prefix,
                              'EXAG' = 'Extensive Agriculture',
                              'INAG' = 'Intensive Agriculture',
                              'PAST' = 'Pastoralism',
                              'HUNT' = 'Foraging',
                              'URBN' = 'Urban',
                              'EXP' = 'Expertise',
                              'DQ' = 'Data Quality'),
         time = time / 1000,
         lower =  plogis(fit - 2 * se),
         upper = plogis(fit + 2 * se),
         fit = plogis(fit))

g1 <- global_trends %>%
  filter(prefix %in% c('HUNT','PAST','EXAG','INAG')) %>%
ggplot(aes(time, fit, group = type)) +
  geom_hline(data = cutpoints_dat, aes(yintercept = plogis(Minimal)), linetype = 2, alpha = .5) +
    geom_hline(data = cutpoints_dat, aes(yintercept = plogis(Common)), linetype = 2, alpha = .5) +
    geom_hline(data = cutpoints_dat, aes(yintercept = plogis(Widespread)), linetype = 2, alpha = .5) +
  geom_line(aes(color = type), size = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax =  upper, fill = type), alpha = .3) +
  scale_color_viridis_d(guide = FALSE) +
  scale_fill_viridis_d(guide = FALSE) +
  scale_x_continuous(breaks = c(-10, -8,-6,-4,-2,0), minor_breaks = NULL) +
  scale_y_continuous(breaks = c(0,1), labels = c('None', 'Widespread'), limits = c(0,1)) +
  labs(x = 'Thousand years BP', y = 'Land-use extent\n') +
  theme_bw() +
    theme(strip.background = element_blank()) +
  facet_wrap(~type)



g2 <- global_trends %>%
  filter(prefix %in% c('EXP', 'DQ')) %>%
ggplot(aes(time, fit, group = type)) +
  geom_line(aes(color = type), size = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax =  upper, fill = type), alpha = .3) +
  scale_color_discrete(guide = FALSE) +
  scale_fill_discrete(guide = FALSE) +
  scale_x_continuous(breaks = c(-10,-8,-6,-4,-2,0), minor_breaks = NULL) +
  scale_y_continuous(breaks = c(0,1), labels = c('None', 'High'), limits = c(0,1)) +
  labs(x = 'Thousand years BP', y = 'Archaeological \nknowledge') +
  theme_bw() +
  theme(strip.background = element_blank()) +
  facet_wrap(~type)

g1 + g2 + plot_layout(ncol = 1, heights = c(2,1)) + plot_annotation(tag_levels = 'A')

ggsave('figures/global_trends.png', height = 6, width = 6.5)
```

```{r, fig.width = 6, fig.height = 4}
global_trends %>%
  filter(prefix %in% c('HUNT','PAST','EXAG','INAG')) %>%
ggplot(aes(time, fit, group = type)) +
  geom_hline(data = cutpoints_dat, aes(yintercept = plogis(Minimal)), linetype = 2, alpha = .5) +
    geom_hline(data = cutpoints_dat, aes(yintercept = plogis(Common)), linetype = 2, alpha = .5) +
    geom_hline(data = cutpoints_dat, aes(yintercept = plogis(Widespread)), linetype = 2, alpha = .5) +
  geom_line(aes(color = type), size = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax =  upper, fill = type), alpha = .3) +
  scale_color_viridis_d(guide = FALSE) +
  scale_fill_viridis_d(guide = FALSE) +
  scale_x_continuous(breaks = c(-10, -8,-6,-4,-2,0), minor_breaks = NULL) +
  scale_y_continuous(breaks = c(0,1), labels = c('None', 'Widespread'), limits = c(0,1)) +
  labs(x = 'Thousand years BP', y = 'Land-use extent\n') +
  theme_bw() +
    theme(strip.background = element_blank()) +
   geom_text(data = cutpoints_dat, aes(-8,plogis(Minimal),label = 'Minimal', vjust = -1)) +
     geom_text(data = cutpoints_dat, aes(-8,plogis(Common),label = 'Common', vjust = -1)) +
   geom_text(data = cutpoints_dat, aes(-8,plogis(Widespread),label = 'Widespread', vjust = -1)) +

  facet_wrap(~type)

cutpoints_dat <- trend_dat %>%
  select(prefix, mod) %>%
  mutate(cutpoints = map(mod, ~.$family$getTheta(TRUE))) %>%
  select(-mod)


c1 <- global_trends %>%
  filter(prefix == 'EXAG') %>%
  ggplot(aes(time, fit, group = type)) +
  geom_hline(yintercept =c(0,plogis(cutpoints_dat[[4,2]])), linetype = 2, alpha = .5) +
  geom_line(aes(color = type), size = 1.2, color = '#377eb8') +
  geom_ribbon(aes(ymin = lower, ymax =  upper, fill = type), alpha = .3, fill = '#377eb8') +
  scale_color_viridis_d(guide = FALSE) +
  scale_fill_viridis_d(guide = FALSE) +
  scale_x_continuous(breaks = c(-10, -8,-6,-4,-2,0), minor_breaks = NULL) +
  scale_y_continuous(breaks = c(0,plogis(cutpoints_dat[[4,2]])), minor_breaks = NULL, labels = c('None', 'Minimal', 'Common', 'Widespread'), limits = c(0,1)) +
  labs(x = '', y = 'Land-use extent\n') +
      theme(strip.background = element_blank()) +
  theme_bw() +
  facet_wrap(~type)

c2 <- global_trends %>%
  filter(prefix == 'INAG') %>%
  ggplot(aes(time, fit, group = type)) +
  geom_hline(yintercept =c(0,plogis(cutpoints_dat[[5,2]])), linetype = 2, alpha = .5) +
  geom_line(aes(color = type), size = 1.2, color = '#4daf4a') +
  geom_ribbon(aes(ymin = lower, ymax =  upper, fill = type), alpha = .3, fill = '#4daf4a') +
  scale_x_continuous(breaks = c(-10, -8,-6,-4,-2,0), minor_breaks = NULL) +
  scale_y_continuous(breaks = c(0,plogis(cutpoints_dat[[5,2]])), minor_breaks = NULL, labels = c('None', 'Minimal', 'Common', 'Widespread'), limits = c(0,1)) +
  labs(x = '', y = '') +
      theme(strip.background = element_blank()) +
  theme_bw() +
  facet_wrap(~type)

c3 <- global_trends %>%
  filter(prefix == 'PAST') %>%
  ggplot(aes(time, fit, group = type)) +
  geom_hline(yintercept =c(0,plogis(cutpoints_dat[[6,2]])), linetype = 2, alpha = .5) +
  geom_line(aes(color = type), size = 1.2, color = '#e41a1c') +
  geom_ribbon(aes(ymin = lower, ymax =  upper, fill = type), alpha = .3, fill = '#e41a1c') +
  scale_color_viridis_d(guide = FALSE) +
  scale_fill_viridis_d(guide = FALSE) +
  scale_x_continuous(breaks = c(-10, -8,-6,-4,-2,0), minor_breaks = NULL) +
  scale_y_continuous(breaks = c(0,plogis(cutpoints_dat[[6,2]])), minor_breaks = NULL, labels = c('None', 'Minimal', 'Common', 'Widespread'), limits = c(0,1)) +
  labs(x = 'Thousand years BP', y = 'Land-use extent\n') +
      theme(strip.background = element_blank()) +
  theme_bw() +
  facet_wrap(~type)

c4 <- global_trends %>%
  filter(prefix == 'HUNT') %>%
  ggplot(aes(time, fit, group = type)) +
  geom_hline(yintercept =c(0,plogis(cutpoints_dat[[3,2]])), linetype = 2, alpha = .5) +
  geom_line(aes(color = type), size = 1.2, color = '#ff7f00') +
  geom_ribbon(aes(ymin = lower, ymax =  upper, fill = type), alpha = .3, fill = '#ff7f00') +
  scale_color_viridis_d(guide = FALSE) +
  scale_fill_viridis_d(guide = FALSE) +
  scale_x_continuous(breaks = c(-10, -8,-6,-4,-2,0), minor_breaks = NULL) +
  scale_y_continuous(breaks = c(0,plogis(cutpoints_dat[[3,2]])), minor_breaks = NULL, labels = c('None', 'Minimal', 'Common', 'Widespread'), limits = c(0,1)) +
  labs(x = 'Thousand years BP', y = '') +
      theme(strip.background = element_blank()) +
  theme_bw() +
  facet_wrap(~type)


c1 + c2 + c3 + c4
```

```{r}
cumsum_landuse_regions <- consensus %>% 
  select(Region, FHG_10KBP:URBAN_1850CE) %>% 
  gather(variable, value, -Region) %>% 
  filter(value %in% c('Widespread', 'Common', 'Split', 'Present')) %>% 
  separate(variable, into = c('land_use_category', 'years_BP'), sep = '_') %>% 
  mutate(land_use_category = ifelse(str_detect(land_use_category, 'AGR'),
                                    str_replace_all(land_use_category , 
                                                    'AGR', 
                                                    'AG'),
                                    land_use_category)) %>% 
  mutate(land_use_category = case_when(
    land_use_category == 'FHG'   ~  'Foraging',
    land_use_category == 'EXAG'  ~  'Extensive Agriculture',
    land_use_category == 'INAG'  ~  'Intensive Agriculture',
    land_use_category == 'PAS'   ~  'Pastoralism',
    land_use_category == 'URBAN' ~  'Urban Centers') 
    ) %>% 
  mutate(years_BP = ifelse(str_detect(tolower(years_BP), 'kbp'), 
                           -parse_number(years_BP) * 1000,
                           ifelse(str_detect(tolower(years_BP), 'ce'), 
                           parse_number(years_BP),
                           -parse_number(years_BP)))) %>% 
  unite(land_use_category_consensus_assessments, 
        c('land_use_category', 
          'value'),
        sep = ' ') %>%
  complete(land_use_category_consensus_assessments,
           nesting(years_BP)) %>% 
  group_by(land_use_category_consensus_assessments, 
           years_BP) %>% 
  summarise(n = n()) %>% 
  mutate(perc = n / sum(n) * 100)  %>% 
  ungroup() %>% 
  # make an ordered factor for nice plotting
  mutate(land_use_category_consensus_assessments = 
           fct_relevel(land_use_category_consensus_assessments, 
                       c('Extensive Agriculture Common',
                         'Extensive Agriculture Widespread',
                         'Intensive Agriculture Common',
                         'Intensive Agriculture Widespread',
                         'Pastoralism Common',
                         'Pastoralism Widespread',
                          'Foraging Common',
                         'Foraging Widespread',
                         'Urban Centers Split',
                         'Urban Centers Present'
                         ))) %>%
  mutate(years_BP = if_else(years_BP > 0, years_BP - 1950, years_BP),
         years_BP = years_BP / 1000)
```

```{r, eval = F, echo = F, fig.width =6.5, fig.height = 4}
ggplot(cumsum_landuse_regions, 
       aes(years_BP, perc, fill = land_use_category_consensus_assessments)) +
  geom_area(position = 'fill', color = 'black') +
  scale_fill_brewer(palette = 'Paired') +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
legend.text = element_text(size = 10),
    legend.key.size = unit(0.7, 'line'),
    legend.position = 'bottom') +
    guides(fill = guide_legend(title = 'Land use based on \nconsensus assessments', nrow = 5, byrow = TRUE)) +
  labs(x = 'Years BP', y = 'Proportion of global land use')
```

```{r, fig.height = 7.5, fig.width = 6.5}
trend_colors <- c('#377eb8', '#4daf4a', '#e41a1c', '#ff7f00')

g1 <- global_trends %>%
  filter(prefix %in% c('HUNT','PAST','EXAG','INAG')) %>%
ggplot(aes(time, fit, group = type)) +
  geom_line(aes(color = type), size = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax =  upper, fill = type), alpha = .3) +
  scale_color_manual(guide = FALSE, values = trend_colors) +
  scale_fill_manual(guide = FALSE, values = trend_colors) +
  scale_x_continuous(breaks = c(-10, -8,-6,-4,-2,0), minor_breaks = NULL) +
  scale_y_continuous(breaks = c(0,1), labels = c('None', 'Widespread'), limits = c(0,1)) +
  labs(x = 'Thousand years BP', y = 'Land-use extent\n') +
  theme_minimal() +
  theme(strip.background = element_blank()) +
  facet_wrap(~type)

g2 <- cumulative_trends <- ggplot(cumsum_landuse_regions, 
       aes(years_BP, perc, fill = land_use_category_consensus_assessments)) +
  geom_area(position = 'stack', color = 'black') +
  scale_fill_brewer(palette = 'Paired') +
  theme_minimal() +
  scale_x_continuous(breaks = c(-10, -8,-6,-4,-2,0)) +
  theme(panel.grid.minor = element_blank(),
legend.text = element_text(size = 10),
    legend.key.size = unit(0.7, 'line'),
    legend.position = 'bottom') +
    guides(fill = guide_legend(title = '', nrow = 5, byrow = TRUE)) +
  labs(x = 'Thousand years BP', y = 'Percent of all regions')

(g1 + theme_minimal()) / cumulative_trends + plot_annotation(tag_levels = 'A')

ggsave('figures/global_trends.png', height = 7.5, width = 6.5)
```


## Regional Trends

```{r, echo = FALSE}
plot_trends <- function(x, variable, color_theme = 'PuOr', tag){
  p1 <- x$trends[[1]] %>%
    gather(time, value, '-10000':'-150') %>%
    ggplot(aes(as.numeric(time) / 1000, value, 
               group = REGION_LAB, color = as.factor(cluster))) +
    geom_line() +
    scale_color_brewer(palette = color_theme, guide = 'none') +
    scale_x_continuous(breaks = c(-8,-6,-4,-2), minor_breaks = NULL) +
    scale_y_continuous(breaks = c(0,1), labels = c('None', 'Widespread')) +
    facet_wrap(~cluster, nrow = 1) +
    labs(x = 'Thousand years BP', y = '') +
    theme_minimal() +
    theme(strip.background = element_blank(),strip.text.x = element_blank())
  
  p2 <- x$trends[[1]] %>%
    select(REGION_LAB, cluster) %>% # just select the columns of interest
    left_join(select(x$data[[1]], REGION_ID:REGION_LAB)) %>%
    group_by(REGION_ID) %>%
    filter(row_number() == 1) %>%
    left_join(regions, ., by = c('Archaeo_ID' = 'REGION_ID')) %>% # join to the region shapes
    mutate(region_label = replace(Archaeo_RG, !(Archaeo_RG %in% c('Hawaii','Polynesia','Micronesia','Melanesia')), NA)) %>%
    ggplot() + # plot
    geom_sf(aes(fill = as.factor(cluster)), size = .3, color = 'white') +
    geom_sf_text(aes(label = region_label), nudge_x = 500000, size = 2.5, hjust = 0) +
    labs(subtitle = variable, x = '', y='', tag = tag) +
    scale_fill_brewer(palette = color_theme, guide = 'none') +  
    coord_sf(datum = NA) +
    theme_void()
  
  # the patchwork command that prints the multipanel plots
  (p2/p1) + plot_layout(heights = c(5,1), widths = c(4,1))
}
```

```{r, fig.width = 12, fig.height=8}
r1 <- trend_dat[4, ] %>%
  mutate(trends = map(trends, ~mutate(.,
                                 cluster = recode_factor(cluster, 
                                 `4` = '1', `2` = '2',`1` = '3', 
                                 `5` = '4', `3` = '5', `6` = '6')))) %>%
  plot_trends('Extensive Agriculture', tag = 'A')

r2 <- trend_dat[5, ] %>%
  mutate(trends = map(trends, ~mutate(.,
                                 cluster = recode_factor(cluster, 
                                `6` = '1', `5` = '2', `3` = '3', 
                                 `1` = '4', `2` = '5', `4` = '6')))) %>%
plot_trends('Intensive Agriculture', tag = 'B')

r3 <- trend_dat[6, ] %>%
  mutate(trends = map(trends, ~mutate(.,
                                 cluster = recode_factor(cluster, 
                                 `3` = '1', `5` = '2',`1` = '3', 
                                 `6` = '4', `4` = '5', `2` = '6')))) %>%
plot_trends('Pastoralism', tag = 'C')

r4 <- trend_dat[3, ] %>%
  mutate(trends = map(trends, ~mutate(.,
                                 cluster = recode_factor(cluster, 
                                 `6` = '1', `5` = '2',`1` = '3', 
                                 `2` = '4', `3` = '5', `4` = '6')))) %>%
  plot_trends('Foraging', tag = 'D')

(r1 | r2 ) / (r3 | r4)

ggsave('figures/trends_map_landuse.png', height = 8, width = 12)
```

```{r}
#temporarily stop knitting here
knitr::knit_exit()
```


```{r echo = FALSE, fig.width = 12, fig.height = 8}
r4
ggsave('figures/trends_foraging.png', height = 8, width = 12)
```

```{r echo = FALSE, fig.width = 6.5}
r3
ggsave('figures/trends_pastoralism.png', height = 8, width = 12)
```

```{r echo = FALSE, fig.width = 6.5}
r1
ggsave('figures/trends_extensive.png', height = 8, width = 12)
```

```{r echo = FALSE, fig.width = 6.5}
r2
ggsave('figures/trends_intensive.png', height = 8, width = 12)
```

```{r echo = FALSE, fig.width = 6.5}
trend_dat[7, ] %>%
  mutate(trends = map(trends, ~mutate(.,
                                 cluster = recode_factor(cluster, 
                                 `2` = '1', `6` = '2',`5` = '3', 
                                 `4` = '4', `1` = '5', `3` = '6')))) %>%
plot_trends('Urbanism')
ggsave('figures/trends_urbanism.png', height = 8, width = 12)
```

```{r, echo = FALSE, fig.width = 9, fig.height = 10}
k1 <- trend_dat[1, ] %>%
  mutate(trends = map(trends, 
                      ~mutate(., cluster = recode_factor(cluster, 
                                 `1` = '1', `2` = '2',`3` = '3', 
                                 `5` = '4', `4` = '5', `6` = '6')))) %>%
  plot_trends('Expertise', color_theme = 'RdYlBu', tag = 'A') + 
  scale_y_continuous(breaks = c(0,1), labels = c('None', 'High')) +
  labs(x = 'Thousand years BP', y = 'Expertise')

k2 <- trend_dat[2, ] %>%
  mutate(trends = map(trends, ~mutate(., 
                                  cluster = recode_factor(cluster, 
                                 `3` = '1', `2` = '2',`6` = '3', 
                                 `1` = '4', `5` = '5', `4` = '6')))) %>%
  plot_trends('Data Quality', color_theme = 'RdYlBu', tag = 'B') + 
  scale_y_continuous(breaks = c(0, 1), labels = c('Unknown', 'Good')) +
  labs( x = 'Thousand years BP', y = 'Quality')

global_trends %>%
  filter(prefix %in% c('EXP', 'DQ')) %>%
ggplot(aes(time, fit, group = type)) +
  geom_line(aes(color = type), size = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax =  upper, fill = type), alpha = .3) +
  scale_color_discrete(guide = FALSE) +
  scale_fill_discrete(guide = FALSE) +
  scale_x_continuous(breaks = c(-10,-8,-6,-4,-2,0), minor_breaks = NULL) +
  scale_y_continuous(breaks = c(0,1), labels = c('None', 'High'), limits = c(0,1)) +
  labs(x = 'Thousand years BP', y = 'Archaeological \nknowledge') +
  theme_bw() +
  theme(strip.background = element_blank()) +
  facet_wrap(~type)

k1 / k2 / g2 + plot_layout(heights = c(5,1,5,1,3))

k1 / k2 + plot_layout()
archaeoglobe
k1;k2
ggsave('figures/trends_exp.png', height = 8, width = 12)
```

```{r, echo = FALSE, fig.width = 6.5}
trend_dat[2, ] %>%
  mutate(trends = map(trends, ~mutate(., 
                                  cluster = recode_factor(cluster, 
                                 `3` = '1', `2` = '2',`6` = '3', 
                                 `1` = '4', `5` = '5', `4` = '6')))) %>%
  plot_trends('Data Quality', color_theme = 'RdYlBu', title = 'Archaeological trends') + 
  scale_y_continuous(breaks = c(0, 1), labels = c('Unknown', 'Good')) +
  labs( x = 'Thousand years BP', y = 'Quality')

ggsave('figures/trends_quality.png', height = 8, width = 12)
```


## HYDE Comparison

```{r}
readxl::read_excel('../hyde_compare_2018_10_26.xlsx')

```

```{r}
inag_onset <- consensus %>%
  select(Region, Label, INAG_10KBP:INAG_1850CE) %>%
  gather(time, value, INAG_10KBP:INAG_1850CE) %>%
  filter(value %in% c('Common', 'Widespread')) %>%
  mutate(time = parse_number(time),
         time = case_when(time == 1500 ~ .5,
                          time == 1750 ~ .25,
                          time == 1850 ~ .15,
                          time <= 10 ~ time),
         time = time * -1) %>%
  group_by(Region, value) %>%
  summarise(onset = min(time)) %>%
  spread(value, onset) %>%
  mutate(Common = if_else(is.na(Common), Widespread, Common))
  
inag_onset %>%
  left_join(regions, ., by = c('Archaeo_ID' = 'Region')) %>%
  mutate(region_label = replace(Archaeo_RG, !(Archaeo_RG %in% c('Hawaii','Polynesia','Micronesia','Melanesia')), NA)) %>%
  ggplot() +
  geom_sf(aes(fill = Common), size = .3, color = 'white') +
  scale_fill_viridis_c(name = 'Onset', guide = 'colorbar', limits = c(-8, -.1)) +
  geom_sf_text(aes(label = region_label), nudge_x = 500000, size = 2.5, hjust = 0, expand = TRUE) +
  coord_sf(datum = NA) +
  theme_void() +
  theme(legend.position='bottom')

inag_onset %>%
  left_join(regions, ., by = c('Archaeo_ID' = 'Region')) %>%
  mutate(region_label = replace(Archaeo_RG, !(Archaeo_RG %in% c('Hawaii','Polynesia','Micronesia','Melanesia')), NA)) %>%
  ggplot() +
  geom_sf(aes(fill = Widespread), size = .3, color = 'white') +
  scale_fill_viridis_c(name = 'Onset (Thousand years BP)', guide = 'colorbar', limits = c(-8, -.1)) +
  geom_sf_text(aes(label = region_label), nudge_x = 500000, size = 2.5, hjust = 0, expand = TRUE) +
  coord_sf(datum = NA) +
  theme_void() 
```
